---
title: "Day 1: Linear Regression"
subtitle: "Molecular Medicine - Oncology, Biostatistics"
author: "Jeroen D. Mulder, Pär Villner, Karl Oskar Ekvall"
date: now
---

In these lab exercises you will get acquinted with performing a simple linear regression analysis in `R`. For this, we use data from $n = 1453$ elderly people who underwent surgery for leg fractures. We recorded baseline information, and contacted them one year later, asking them to what degree they can move their broken leg recorded on a scale of 0 to 100 (`y1`), and if they still have pain with `0 = no` and `1 = yes` (`y2`). 

You will look at three examples:

- The relationship between the movability of the broken leg (`y1`) and the body mass index (`bmi`);
- The relationship between the moveabilty of the broken leg (`y1`) and the type of `city` (small, medium, or large); and 
- The relationship between the body mass index (`bmi`) and `age`. 

You will run multiple analyses using the `lm()` function. Make sure you understand the following in the `summary()` output of the `lm()` command:

- Interpretation of parameter estimates,
- Confidence intervals for parameters,
- t-values,
- p-values, and 
- $R^2$ value.

# Data Description
Load the data into R using the `read.csv()` function: 

```{r setup}
#| echo: true
#| output: false
# Read in data
d1 <- read.csv(
  "Biostatsweek_data1.csv",
  sep = ";",
  header = TRUE,
  stringsAsFactors = TRUE,
  dec = ","
)

# Inspect the first few rows of data
head(d1)
```

In addition to the variables menionted above, it contains: 

- `hosp` = hospital in which they were treated
- `severity` = severity of the fracture (1-10)
- `sex` (m/f)
- `edu` = education (low/medium/high)
- `smoke` = smoking habits (0/1)
- alcohol consumption (dl/day of alcoholic beverages)
- `milk` = milk consumption (dl/day)
- the existance of comorbidities (`ost` = osteoporosis; `cvd`; `cancer`; `diab` = diabetes) (0/1)
- `partner` = whether they have a partner (0/1)
- `city` = whether they live in a small/medium/big town

# Assumptions
Always, before running analyses, it is important to get a "aquinted with"/"get a feeling for" the data. Answer the below questions to check whether assumptions for linear regression hold.

### Question 1

Make histograms of `bmi` and `y1`. Based on the histograms, are there any signs that a linear regression model with `y1` as outcome and `bmi` as independent variable is not suitable?

::: {.callout-note appearance="simple" collapse="true" title="Answer"}

Looking at histograms like this is not a good way of seeing if linear regression is suitable. Still, no obvious warning signs (e.g., no obvious indications that there are non-linear associations).

```{r}
# Visualize distribution of bmi using a histogram
hist(d1$bmi)

# Visualize distribution of y1 using a histogram
hist(d1$y1)
```
:::

### Question 2

Create a scatterplot of `bmi` and `y1`. Does it look like a linear regression model is suitable for this data?

::: {.callout-note appearance="simple" collapse="true" title="Answer"}

Looks like we have a negative and approximately linear association, so a linear regression model would be fine.

```{r}
# Visualize relationship between bmi and y1 using a scatterplot
plot(d1$bmi, d1$y1)
```

The assumption of equal variances can best be checked after the model is fitted, and inspecting the residuals. 
:::

# Moveability and Body Mass Index
Fit a linear regression model which says that the mean `bmi` depends on `y1` using the `lm()` function. That is, we try to predict `bmi` from the movability of the broken leg (`y1`).^[Whether or not this is a sensible question to ask/sensible model to fit is a different matter.] Use the `summary()` function to get more information about the model. Look at the different sections (and their headers) in the output. Then, interpret the results. 

### Question 3
Explain how the $t$-value for the parameter of `y1` is calculated and how the $t$ value can be used to calculate the $p$-value. Run `confint()` and explain how the output of the `summary()` function can be used to calculate the confidence intervals.

::: {.callout-tip}
## Tip
See slide 27 from the lecture. 
:::

::: {.callout-note appearance="simple" collapse="true" title="Answer"}

First, `y1` has a slight negative effect on the expected `bmi`. The result is statistically significant (the p-value is less than our significance level of 0.05).

```{r}
# Fit a linear regression model
mod1 <- lm(bmi ~ y1, data = d1)

# Print a summary of the results
summary(mod1)
```

Second, the $t$-value is the estimate divided by its standard error; the $p$-value is obtained from the $t$-distribution with `n - 2` degrees of freedom. 

```{r}
#| echo: true
#| output: false
# Save the summary output
s1 <- summary(mod1)

# Print the summary output
s1

# Get the estimate of the parameter of y1
est_y1 <- s1$coefficients[2, 1]

# Get the standard error of the parameter of y1
se_y1 <- s1$coefficients[2, 2]

# Calculate the t-value
t_y1 <- est_y1 / se_y1
t_y1

# Calculate the p-value
p_y1 <- 2 * pt(abs(t_y1), nrow(d1) - 2, lower.tail = FALSE)
p_y1
```

Third, confidence intervals are build using both the estimate and its uncertainty. It is computed as estimate ± (standard error × critical `t`-value).

```{r}
#| echo: true
#| output: false
# Manually compute the 95% confidence interval using summary output
s1$coefficients[1, 1] - s1$coefficients[1, 2] * qt(0.975, nrow(d1) - 2)
s1$coefficients[1, 1] + s1$coefficients[1, 2] * qt(0.975, nrow(d1) - 2)

# Manually compute the 95% confidence interval using the printed numbers (which also come from summary output)
28.58942389 - 0.274581637 * qt(0.975, nrow(d1) - 2)
28.58942389 + 0.274581637 * qt(0.975, nrow(d1) - 2)

# Shortcut method in R
confint(lm(bmi ~ y1, data = d1))
```
:::

### Question 4

Calculate the correlation between `bmi` and `y1`. Is it consistent with the result from Question 3? Why? How?

::: {.callout-note appearance="simple" collapse="true" title="Answer"}

The correlation has the same sign as the coefficient. The parameter estimate can be written as a function of the correlation and standard deviations: 

$$
\beta_1 = \frac{Cor(X, Y)}{\sqrt{Var(X) \cdot Var(Y)}}
$$

```{r}
#| echo: true
#| output: false
# Calculate the correlation between bmi and y1
cor(d1$bmi, d1$y1)

# Compute the regression coefficient using the correlation and standard deviations
beta1 <- cor(d1$bmi, d1$y1) * sd(d1$bmi) / sd(d1$y1)
beta1
```
:::

### Question 5

Plot a histogram of the residuals from the regression model in Question 3. Does the distribution of the residuals seem approximately normal? Why does it (not) matter?

::: {.callout-note appearance="simple" collapse="true" title="Answer"}

Looks approximately normal, but thanks to the central limit theorem the distribution of our residuals is not very important for us, as long as the sample size is large enough.

```{r}
#| echo: true
#| output: true
# Obtain residuals from summary output
residuals <- s1$residuals

# Plot a histogram of the residuals
hist(residuals)
```
:::

### Question 6

Now run the same analysis as in Question 3 but now treat mean `y1` as the outcome and `bmi` as the regressor (independent variable). That is, we ask the question whether the movability of the broken leg (`y1`) can be predicted by the body mass index (`bmi`) of a patient. Interpret the parameter estimates.

::: {.callout-note appearance="simple" collapse="true" title="Answer"}

There is a negative association between `y1` and `bmi`: The expected `y1` goes down by 2.79 per unit increase in `bmi`.

```{r}
mod2 <- lm(y1 ~ bmi, data = d1)
summary(mod2)
```
:::

### Question 7

Use the model from Question 7 to estimate the expected `y1` of people with the following `bmi`: 5, 25 and 40. Are there reasons to be skeptical of any of these estimates?

::: {.callout-note appearance="simple" collapse="true" title="Answer"}

Yes, we extrapolate a lot with `bmi` = 5 and 40, and the estimates are unreasonable, in particular for `bmi` = 5. In other words, the data do not contain information about `bmi` values of 5 or 40. Hence, predictions for these values are not trustworthy.

```{r}
#| echo: true
#| output: false
# Get the coefficients from the model
b1hat <- mod2$coefficients[1]
b2hat <- mod2$coefficients[2]

# Compute the expected y1 for different bmi values
b1hat + b2hat * 5
b1hat + b2hat * 25
b1hat + b2hat * 40

# As a shortcut, we can fill in a formula with multiple predictor values at once
b1hat + b2hat * c(5, 25, 40)

# Alternatively, we can use the predict() function 
predict(mod2, newdata = data.frame(bmi = c(5, 25, 40)))
```
:::

# City and Moveability

Next, we might want to know if there are differences between cities in terms of the movability of the broken leg (`y1`) of patients. First, graphically illustrate the difference in `y1` distribution between people from big, medium and small cities. Then, fit a linear regression model which in which the mean `y1` depends on the city variable. 

::: {.callout-tip title="Tip"}
Think about what kind of variable the city variable is (categorical or continuous). How does this change the analysis? 
:::

### Question 8

What is the difference in expected mean `y1` between people from small and medium cities according to the model?

::: {.callout-note appearance="simple" collapse="true" title="Answer"}

The city variable is categorical, containing three levels: Big, medium and small. Hence, fitting a linear regression model with `city` as a predictor requires the use of a dummy variable. Luckily, the `lm()` function in R can handle this automatically when we indicate that the city variable is a factor. For this, we can use the `factor()` function.

```{r}
#| echo: true
#| output: false
# Graphically illustrate the difference in y1 distribution between people from big, medium and small cities
boxplot(d1$y1 ~ d1$city)

# Visually, the differeces are not very large.

# Fit a linear regression model with city as a predictor
mod3 <- lm(y1 ~ factor(city), data = d1)
summary(mod3)
```
Once you have the coefficients, you can compute the difference in expected mean `y1` between people from small and medium cities. For this, it is easiest to write down the full regression equation, and then fill it in for the two city types.

$$
\begin{aligned}
\mathbb{E}[y1 | city] &= \beta_0 + \beta_1 \cdot Medium + \beta_2 \cdot Small\\
\\
\mathbb{E}[y1 | city = \text{small}] &= \beta_0 + \beta_1 \cdot 0 + \beta_2 \cdot 1 \\
\mathbb{E}[y1 | city = \text{medium}] &= \beta_0 + \beta_1 \cdot 1 + \beta_2 \cdot 0 \\
\mathbb{E}[y1 | city = \text{big}] &= \beta_0 + \beta_1 \cdot 0 + \beta_2 \cdot 0 \\
\end{aligned}
$$
Then, filling in the estimated coefficients in the equation for small and medium cities, we get: 
```{r}
# Small cities
y1_hat_small <- mod3$coefficients[1] + mod3$coefficients[2] * 0 + mod3$coefficients[3] * 1
y1_hat_small

# Medium cities
y1_hat_medium <- mod3$coefficients[1] + mod3$coefficients[2] * 1 + mod3$coefficients[3] * 0
y1_hat_medium

# Difference is
y1_hat_medium - y1_hat_small
```
:::

### Question 9

Do the residuals from the model in Question 8 look normally distributed? Is there a relationship between the residuals and the city variable?

::: {.callout-note appearance="simple" collapse="true" title="Answer"}

Residuals look slightly negatively skewed, but this is not a big problem given the large sample size (central limit theorem). The distribution of the residuals seems to be the same across all city types (assumption of equal variances across level of the predictor variable).

```{r}
hist(summary(mod3)$residuals)

boxplot(summary(mod3)$residuals ~ d1$city)
```
:::

### Question 10

Test the hypothesis that the mean `y1` is the same for people from big, medium and small cities. How do you write down such a hypothesis? Are there any reasons not to trust the test result?

::: {.callout-note appearance="simple" collapse="true" title="Answer"}
The hypothesis is that the mean `y1` is the same for people from big, medium and small cities. We can write this down as:
$$
H_0: \mu_{small} = \mu_{medium} = \mu_{big}
$$
$$
H_a: \mu_{small} \neq \mu_{medium} \neq \mu_{big}
$$

To test this hypothesis with one single test, we can use an ANOVA test, rather than comparing the means of the three city types pairwise using t-tests. An ANOVA can be run in R using the `aov()` function.

```{r}
#| echo: true
#| output: false
# Run an ANOVA test
aov_test <- aov(y1 ~ city, data = d1)

# Print the summary of the ANOVA test
summary(aov_test)
```

The p-value is 0.119, so we do not reject the null hypothesis. In other words, we do not have evidence to suggest that the mean `y1` is different for people from big, medium and small cities. This conclusion corresponds with our visualization of the data.
:::

# Body Mass Index and Age
Fit a linear regression model which explains how mean `bmi` depends on `age`. 

### Question 11
Interpret the result and argue whether or not we should trust the model.

::: {.callout-note appearance="simple" collapse="true" title="Answer"}

With a term such as "trust the model", we refer to wheter or not the assumptions underlying an analysis are met. We can test the assumption of linearity with a scatterplot of the raw data, and the assumption of equal variances with a boxplot of the residuals. The assumption of independence of observations is not necessarily testable with data, and must be reasoned about based on the study design.

The expected `bmi` for a person of `age = 0` is 23.5, but this is a nonsensical estimate. Age hardly seems to matter for `bmi`: An increase in age by 1 year is associated with a decrease in expected `bmi` by 0.01201, which is not significantly different from 0, $p = .369$. The model seems trustworthy based on the visualizations.

```{r}
#| echo: true
#| output: false
# Fit a linear regression model
mod4 <- lm(d1$bmi ~ d1$age)

# Print the summary of the model
summary(mod4)

# Visualize the relationship between bmi and age using a scatterplot
plot(d1$bmi ~ d1$age)

# Plot the fitted line through the data
abline(mod4$coef, col = "red")
# A straight (linear) line appears to fit the data well

# Visualize the relationship between age and residuals using a scatterplot
plot(d1$age, residuals(mod4))
# The residuals are equally scattered around across the range of age, indicating that the assumption of equal variances is met.
```
:::

# Cancer and Body Mass Index
Fit a linear regression model which explains how mean `bmi` depends on cancer status. 

### Question 12
Interpret the result and argue whether or not we should trust the model. Also test the hypothesis that cancer patients have the same mean `bmi` as non-cancer patients.

::: {.callout-note appearance="simple" collapse="true" title="Answer"}

According to the model, cancer patients have a slightly higher expected `bmi` (0.23 higher than non-cancer patients). But the difference in mean values is not statistically significant on any standard significance level, as $p = .21$. This result is not surprising based on the visualizations of the data. The model looks trustworthy based on visualization of data/residuals.

To test the hypothesis that the means are equal, it would be enough to look at the p-value for the cancer parameter in our model, but we can also perform a t-test to reach the same conclusion.

```{r}
#| echo: true
#| output: false
# Visualize the relationship between bmi and cancer status using a boxplot
boxplot(d1$bmi ~ d1$cancer)

# Fit a linear regression model with cancer status as a predictor
mod5 <- lm(bmi ~ factor(cancer), data = d1)
summary(mod5)

# Visualize the relationship between cancer status and residuals using a boxplot
boxplot(residuals(mod5) ~ factor(d1$cancer))

# Perform a t-test to test the hypothesis that cancer patients have the same mean `bmi` as non-cancer patients
t.test(d1$bmi ~ factor(d1$cancer))
```
:::
