---
title: "Day 2: Logistic Regression"
subtitle: "Molecular Medicine - Oncology, Biostatistics"
author: "Jeroen D. Mulder, Pär Villner, Karl Oskar Ekvall"
date: now
---

In these lab exercises you will gain experience with logistic regression as an analysis technique, and learn how to perform this method in `R`. For this, we use data from 376 patients with colorectal cancer with whom we measured carcinoembryonic antigen (CEA), a tumor biomarker. The goal of these exercises is twofold: 

1. Study how the biomarker levels depend on the stage of the disease.
2. Determine how CEA can predict 1-year survival.

# Data Description and Setup
The data include the following variables: 

- `age` and `sex` of each patient; 
- the `stage` of the disease (an integer between 1 and 5); 
- `CEA`, expressed in g/L; 
- a binary indicator `dead_1yr`, which is 1 if the patient is dead by 1 year from the data collection, and zero otherwise.

Load the data into R using the `readRDS()` function: 
```{r setup}
#| echo: true
#| output: false
d2 <- readRDS("data2.Rds")
str(d2) # Should show data.frame with 376 obs of 5 variables
```

Moreover, below we will occasionally use the logistic function. Hence, it is useful to define it in `R` such that we can easily call it when it is needed. Here, `a` is the name for the linear prediction, that is $a = \beta_{0} + \beta_{1} x$: 

```{r logistic}
#| echo: true
#| output: false
# Define the logistic function
g <- function(a) { (1 + exp(-a))^(-1) }
```

# Descriptive Analyses
Before fitting a logistic regression model, we first try to get a better "feeling" for the data by performing some descriptive analyses. 

### Question 1

Create a contingency table of the disease stage `stage` and 1-year survival `dead_1yr`. Interpret the table. Does it look like there is an association between disease stage and 1-year survival?

::: {.callout-tip}
## Tip
Think about which pattern (in terms of proportion) you would expect if there was no relation between disease stage and 1-year surival. In that case, across all stages of disease, the proportions of patients who died within one year would be .... Is this also the type of relation that you see in the data? 
:::

::: {.callout-note appearance="simple" collapse="true" icon=false title="Answer"}

The proportion of patients who died within one year clearly changes (in favor of dying) as the disease stage progresses. Hence, it looks like there is an association between these to variables. In particular, stage 1 and 2 are associated with lower risks than the higher stages.

```{r}
#| echo: true
#| output: false
# Create a cross-table displaying counts of combinations of factor levels
tab <- table(d2$stage, d2$dead_1yr)

# Transform the counts of a cross-table to proportions
prop.table(tab, margin = 1) # `margin = 1` indicates proportions by row
```
:::

### Question 2

Calculate the odds that a randomly selected patient with stage 1 dies within a year, and the odds that a randomly selected patient with stage 4 dies within a year. Then use the two odds to calculate an odds ratio. Explain in words what this odds ratio means.

::: {.callout-note appearance="simple" collapse="true" icon=false title="Answer"}
The odds of dying within one year are computed as $\frac{P(Y = dead)}{1 - P(Y = dead)}$. Hence, we first need to compute the probabilities $P()$. Second, we can use the probabilities to compute the odds. These calculations need to be done for the stage 1 and stage 4 patients seperately. 

```{r}
#| echo: true
#| output: false
# Compute the probabilities for the stage 1 and stage 4 patients
p1 <- tab[1, 2] / sum(tab[1, ])
p4 <- tab[4, 2] / sum(tab[4, ])

# Compute the odds
o1 <- p1 / (1 - p1)
o4 <- p4 / (1 - p4)

# Compute the odds ratio
o4 / o1
```

Explaining what the odds ratio means in everyday language is difficult, but we can repeat the mathematical definition: It is the ratio of two odds. See slides for more details.
:::

# Logistic Regression
Next, rather than only looking at (nondirectional) associations, we perform logistic regression to study if we can predict 1-year survival based on stage. 

### Question 3

Fit a simple logistic regression model where the response is `dead_1yr` and the predictor is `stage` (treated as factor, not as a numeric). A logistic regression can be run in `R` using the `glm()` function and setting as one its arguments `family = binomial`. Use the `summary()` function and comment on the output. Also calculate the probabilities that persons with stage 4 and stage 1 are dead within a year.

::: {.callout-note appearance="simple" collapse="true" icon=false title="Answer"}
We fit a logistic regression model with the code below. Note the use of the `factor()` function, which indicates to `glm()` that the `stage` variable should be treated as an ordered categorical variable rather than as as continuous predictor. As a consequence, the `stage` variable will be split into dummy variables. 

```{r}
#| echo: true
#| output: false
# Fit a logistic regression model (with dummy variables)
mod1 <- glm(dead_1yr ~ factor(stage), data = d2, family = binomial)
```

All parameters are significant, and the positive parameter estimates indicate a higher probability than 0.5 for people of stage 1 to die within a year, as well as an increased risk of death for people with higher stage values. The pattern from the frequency and proportion table is here as well.

```{r}
#| echo: true
#| output: false
# Summarize model output
summary(mod1)

# Interpret model coefficients using the logistic function: 
g(mod1$coefficients[1]) # Predicted probability for stage 1
g(mod1$coefficients[1] + mod1$coefficients[4]) # Predicted probability for stage 4

# Compare to estimates based on the table
p4
```

Note how the predictor probabilities from the exploratory descriptive analyses (from Question 1) and those from the logistic regression model are equivalent. This is because the logistic regression model contains as many parameters as there are pieces of information in the data, and hence the model is saturated and fits perfectly. 
:::

### Question 4

Give a 95% confidence interval for the odds ratio of being dead within a year for stage 4 relative to stage 1.

::: {.callout-note appearance="simple" collapse="true" icon=false title="Answer"}
We can simply request confidence intervals from (most) fitted models in `R` using the `confint()` function. Note, to ensure the parameters are the odds ratio scale, we still need to exponentiate using `exp()`. 

```{r}
#| echo: true
#| output: false
exp(confint(mod1)[4, ])
```
:::

### Question 5

Rerun the logistic regression model, but now treat `stage` as a numeric. Use the estimates from this model to calculate the probability that persons with stage 4 and stage 1 are dead within a year.

::: {.callout-note appearance="simple" collapse="true" icon=false title="Answer"}
One can explicitely tell the `glm()` function that `stage` is a continuous variable by wrapping it in `as.numeric()`. Consequently, *no* dummy variables are created and hence fewer parameters are estimated. 

```{r}
#| echo: true
#| output: false
# Fit a logistic regression (with continuous predictor)
mod2 <- glm(dead_1yr ~ as.numeric(stage), data = d2, family = binomial)

# Summarize model output
summary(mod2)

# Compute predicted probabilities for stage 1 and stage 4
g(mod2$coefficients[1] + 4 * mod2$coefficients[2]) # Stage 4
g(mod2$coefficients[1] + 1 * mod2$coefficients[2]) # Stage 1 
```
:::

### Question 6

Why are the probability estimates different depending on whether stage is treated as a factor or as a numerical? Which method is preferable?

::: {.callout-note appearance="simple" collapse="true" icon=false title="Answer"}
If we treat `stage` as a numerical variable, we are stating that *each* one-unit increase in disease stage leads to a $e^{\beta_{1}}$ multiplicative change in the odds, whether we go from stage 1 to stage 2, stage 2 to stage 3, etc. However, the preliminary inspection of the contingency table changes that the change in odds was stronger when going from stage 1 to stage 2, then from stage 4 to 5 (for example). Below, we see that we underestimate the risk for people of stage 3 of dying within a year if we use stage as a numerical variable.
```{r}
#| echo: true
#| output: false
# Stage 3
g(mod2$coefficients[1] + 3 * mod2$coefficients[2])
g(mod1$coefficients[1] + mod1$coefficients[3])
# Value based on the contingency table is the same as with the regression model treating stage as a factor variable:
tab[3, 2] / (sum(tab[3, ]))
```
When treating `stage` as an ordered categorical variable, we specify a more complex model (i.e., with more parameters), which is better able to accommodate this patters in the data. However, this is not to say that a model with more parameters is always better. After all, a model is supposed to be a simplified version of reality and hence we should always be balancing model complexity with fit to the data. 
:::

### Question 7

Fit a simple logistic regression model where the response is `dead_1yr` and the predictor is `CEA`. Use `summary()` and comment on the results.

::: {.callout-note appearance="simple" collapse="true" icon=false title="Answer"}

The effect of `CEA` on 1-year survival is estimated to be $\beta_{1} = 0.6230$ which, on the odds ratio scale, is $e^{\beta_{1}} = e^{0.6230} = 1.8645$. Hence, this indicates that increases in `CEA` are associated with an increased risk of dying.

```{r}
#| echo: true
#| output: false
# Inspect the density of the `CEA` predictor
hist(d2$CEA)

# Fit a logistic regression model 
mod3 <- glm(dead_1yr ~ CEA, data = d2, family = binomial)

# Inspect the model output
summary(mod3)
```
:::

### Question 8

Visualize the distribution of `CEA` for participants who died within a year and survived a year. Does it support the estimates of the regression model?

::: {.callout-note appearance="simple" collapse="true" icon=false title="Answer"}

To study this, we can simply split the data by `dead_1yr` and then plot `CEA` for each group. This can be done in multiple steps, are all simultaneously using `boxplot()`. The visualization shows that people who died within one year are having slightly higher CEA values.

```{r}
boxplot(d2$CEA ~ d2$dead_1yr)
```
:::

### Question 9

Calculate the probabilities of being dead within one year for people with CEA levels of 2, 5 and 10. Should we trust these estimates?

::: {.callout-note appearance="simple" collapse="true" icon=false title="Answer"}
To predict the probabilities of being dead within one year for people with CEA levels of 2, 5 and 10, we simply fill in the regression equation again just like in the lab on day 1. 

```{r}
#| echo: true
#| output: false
g(mod3$coefficients[1] + mod3$coefficients[2] * c(2, 5, 10))
```
I would not trust the `CEA = 10` estimate, since no one has a CEA close to this value.
:::

### Question 10

In the below code, I (a) create a logistic regression model with `dead_1yr` as outcome and `age` as regressor, and (b) create a residual plot to check if the model fits the data well. Can you understand what the code does and interpret the plot?

```{r}
#| echo: true
#| output: false
agedead <- glm(dead_1yr ~ age, data = d2, family = binomial)

cbind(agedead$fitted.values, residuals(agedead, type = "response"), d2$dead_1yr, d2$dead_1yr - agedead$fitted.values)

plot(d2$age, residuals(agedead, type = "response"))
scatter.smooth(d2$age, residuals(agedead, type = "response"), family = "gaussian")
```

::: {.callout-note appearance="simple" collapse="true" icon=false title="Answer"}
Look at the comments in the code to check what each line of code does
```{r}
#| echo: true
#| output: false
# Fit a logistic regression model in which 1-year survival is predicted by age
agedead <- glm(dead_1yr ~ age, data = d2, family = binomial)

# Combine various columns in a single data frame
cbind(
  agedead$fitted.values, # The predicted (fitted) values based on the model
  residuals(agedead, type = "response"), # The model residuals
  d2$dead_1yr, # The observed outcomes
  d2$dead_1yr - agedead$fitted.values # The model residuals, manually computed
)

# Plot the residuals against the predictor age
plot(d2$age, residuals(agedead, type = "response"))

# Draw a line through the plot to better see the relation between the residuals and the predictor
scatter.smooth(d2$age, residuals(agedead, type = "response"), family = "gaussian")
```

What makes this graph hard to read is that the circles we see consist of many people overlapping each other. The line shows an average of the residuals. Since it is around zero, it means that out of the participants who were given a particular probability of dying, the same proportion also died. In other words, our model is good, although the fit is not perfect for younger ages (but we have few observations there).
:::
